---
title: "Marital Status and All-Cause Mortality in U.S. Adults: A Survey-Weighted Survival Analysis"
author: "Mike Zhang"
date: today
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

library(tidyverse)
library(haven)
library(nhanesA)
library(gtsummary)
library(survey)
library(survival)
library(survminer)
library(gtsummary)
library(broom)

nhanes_joint <- readRDS(here::here("data/processed/nhanes_joint.rds")) |> 
  mutate(
    surv_object = Surv(permth_int, mortstat)
  )

```

# Background

# Methods

## Data Source

Data for this study were drawn from the National Health and Nutrition Examination Survey (NHANES), a nationally representative survey of the civilian non-institutionalized US population conducted by the Centers for Disease Control and Prevention (CDC). Five consecutive two-year cycles from 2005 to 2014 were pooled to maximize sample size and mortality follow-up. Mortality status was obtained through linkage with the National Death Index (NDI), with follow-up extended through December 2019. All data are publicly available through the CDC National Center for Health Statistics website.

## Study Population

Of the 50,965 participants across five NHANES cycles, those with eligible mortality data and aged 40 years or older were retained for analysis. Participants with missing marital status were subsequently excluded, yielding an analytic sample of `{r} nrow(nhanes_joint)` participants. A further 7,654 participants were excluded from multivariable models due to missing covariate data, resulting in a final analytic sample of 10,950 participants for the Cox proportional hazards analysis. A comparison of included and excluded participants is presented in the supplementary materials.

## Exposure and Outcome

Marital status was assessed via self-report and originally categorized into six groups: married, widowed, divorced, separated, never married, and living with a partner. For analytical purposes, these were collapsed into three categories: married, previously married (widowed, divorced, or separated), and never married (never married or living with a partner). Never married served as the reference category in all models.

The outcome of interest was all-cause mortality. Follow-up time was measured in months from the NHANES interview date to the date of death or censoring. Participants who had not died by December 31, 2019 were censored at that date.

## Covariates

Covariates in the analysis were mainly demographic, behavioral, and clinical. Demographic covariates included were: age in years, sex (male or female), level of education (less than 9th grade, 9-11th grade, high school grad, some college, college grad or above ), ethnicity (Mexican American, Other Hispanic, non-Hispanic white, non-Hispanic black, and other race - including multi-racial), and family-income-to-poverty ratio. Behavioral covariates consisted of smoking status (ever having smoked 100 or more cigarettes) and vigorous physical activity (physically active on a typical day). Clinical covariates included the Body Mass Index (BMI), diabetes history, hypertension history, cardiovascular disease history (either coronary heart disease or myocardial infarction), and self-rated health (poor, fair, good, very good, excellent).

## Statistical Analysis

Because NHANES is a weighted survey, `survey` package was utilized to create a `svydesign` object with accurately identified weights, id, and strata. Un weighted KM curves were shown below, and it demonstrated that the hazards for mortality were different for participants with different marital statuses. A Cox proportional hazards model with all the aforementioned variables was constructed using `svycoxph`. The proportional hazards assumption was tested using `cox.zph()` to examine scaled Schoenfeld residuals of each variable. All variables returned a non-significant p-value of either 0.99 or 1, indicating that the proportional hazards assumption held. Effect modification by sex and age group (40–59, 60–74, 75+) was investigated with by integrating interaction terms between marital status and each modifier in separate models, and both interaction terms were non-significant results. All analyses were conducted using R version 4.6.0.

```{r}
#| label: fig-km
#| fig-cap: "Kaplan-Meier survival curves by marital status among U.S. adults aged 40 and older, NHANES 2005–2014. Curves are unweighted and presented for descriptive purposes."
#| fig-width: 8
#| fig-height: 6

km_fit_unweighted <- survfit(surv_object ~ DMDMARTL, data = nhanes_joint)
ggsurvplot(
  km_fit_unweighted,
  data = nhanes_joint,
  pval = TRUE,
  conf.int = TRUE,
  risk.table = TRUE,
  legend.labs = c("Never married", "Previously married", "Married"),
  legend.title = "Marital Status",
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  title = "Kaplan-Meier Survival Curves by Marital Status",
  ggtheme = theme_minimal()
)
```

# Results

## Sample Characteristics

```{r}
#| label: tbl-1
#| fig-cap: "Baseline characteristics of study participants by marital status, NHANES 2005–2014"
 
nhanes_survey <- svydesign(
  data = nhanes_joint, 
  strata = ~SDMVSTRA, 
  id = ~SDMVPSU, 
  nest = TRUE, 
  weights = ~WTINT2YR
)

nhanes_table <- tbl_svysummary(
  nhanes_survey, 
  by = DMDMARTL,
  include = c(RIDAGEYR, RIAGENDR, DMDEDUC2, INDFMPIR, RIDRETH1, BMXBMI, SMQ020, PAQ605, 
              BPQ020, DIQ010, HSD010, cvd_history),
  label = list(
    RIDAGEYR       ~  "Age (years)",
    RIAGENDR       ~  "Sex",
    DMDEDUC2       ~  "Education",
    INDFMPIR       ~  "Family income-to-poverty ratio",
    RIDRETH1       ~  "Ethnicity",
    BMXBMI         ~  "BMI",
    SMQ020         ~  "Ever smoked 100 cigarettes, Yes",
    PAQ605         ~  "Vigorous physical activity, Yes",
    cvd_history    ~  "Cardiovascular disease history, Yes",
    BPQ020         ~  "Hypertension, Yes",
    DIQ010         ~  "Diabetes, Yes",
    HSD010         ~  "Self-rated health"
  ) 
) |> 
  add_p() |> 
  add_overall()

nhanes_table

```

During a median follow-up of `{r} median(nhanes_joint$permth_int)` months, `{r}  sum(nhanes_joint$mortstat == 1)` deaths occurred among `{r} nrow(nhanes_joint)` eligible participants. After accounting for sample weights, the majority of participants were married (n = 17.3 million), followed by previously married (n = 7.1 million) and never married (n = 3.1 million). Among covariates, previously married participants had the highest burden of hypertension (50%), diabetes (16%), and CVD history (9.7%), the lowest physical activity (14%), and the worst self-rated health distribution. Across marital status groups, the never married group was the youngest. The previously married group had more women than men (67% vs 33%). The married group had the highest family-income-to-poverty ratio (median = 2.11) and the best self-rated health.

## Survival Curves

AKaplan-Meier survival curves stratified by marital status are shown in Figure 1. The previously married group had the worst survival compared to the married and never married groups. Its KM curve separates over time from the other two, indicating higher mortality risk than the other two groups. There isn't any visually apparent difference between survivals of the married and never married groups. However, the KM curves in the plot were unadjusted and unweighted. Formal inference is drawn from the survey-weighted Cox proportional hazards model presented below.

## Cox Proportional Hazards Model

```{r}
cox_fit <- readRDS(here::here("output/models/cox_fit.rds"))

tbl_regression(
  cox_fit,
  exponentiate = TRUE,
  label = list(
    DMDMARTL   ~ "Marital status",
    RIDAGEYR   ~ "Age (years)",
    RIAGENDR   ~ "Sex",
    DMDEDUC2   ~ "Education",
    INDFMPIR   ~ "Income-to-poverty ratio",
    RIDRETH1   ~ "Race/ethnicity",
    BMXBMI     ~ "BMI",
    SMQ020     ~ "Smoking history",
    PAQ605     ~ "Vigorous physical activity",
    cvd_history ~ "CVD history",
    BPQ020     ~ "Hypertension",
    DIQ010     ~ "Diabetes",
    HSD010     ~ "Self-rated health"
  )
)

```

A Cox proportional hazards model was constructed with mortality as the outcome and all covariates as exposures. Married participants had a 42% lower instantaneous mortality rate than never married participants (HR = 0.58, p < 0.001). Notably, while unadjusted Kaplan-Meier curves suggested worse survival among previously married participants, the adjusted Cox model revealed a borderline non-significant reduction in mortality hazard compared to the never married group (HR = 0.80, p = 0.062), suggesting that the unadjusted survival differences were largely attributable to differences in demographic and clinical characteristics across marital status groups. 

As expected, age was a significant predictor of mortality with each additional year of age associated with a 10% higher instantaneous mortality rate. Females had a 36% lower instantaneous mortality rate than males after accounting for all covariates, which was consistent with established literature. Self-rated health showed a strong dose-response effect. Fair and poor health were the only significant levels compared to excellent, with fair health more than doubling the mortality risk and poor health nearly tripling it (HR = 2.12, p < 0.001; HR = 2.9, p < 0.001). Clinical and behavioral covariates performed as expected: hypertension, diabetes, CVD history, and smoking were each associated with increased mortality hazard, while vigorous physical activity was protective.

Tests of effect modification revealed no significant interactions between marital status and sex, with interaction term p-values of 0.68 and 0.59 for previously married and married, respectively. Similarly, no significant interactions were observed between marital status and age group, with all four interaction term p-values exceeding 0.08.

# Limitations

# Conclusions
